### 2024-05-19 Paper/MSEPM-Manuscript.tex
Manuscript revision

\large{\bf{The Multi-State Epigenetic Pacemaker enables the 
identification of combinations of factors that influence DNA methylation}}\\ \\
\noindent{\small Colin Farrell$^{1,3}$, Kalsuda Lapborisuth$^{1}$, 
                 Sagi Snir$^{2}$,\\ and Matteo Pellegrini$^{1,3}$\\ \\
\noindent{\footnotesize 
$^1$Dept. of Molecular, Cell and Developmental Biology; \\ University of
California, Los Angeles, CA 90095, USA;;\\
\noindent
$^2$Dept. of Evolutionary Biology, University of  Haifa, Israel;\\
$^3$Corresponding Authors; colinpatfarrell@g.ucla.edu, matteop@mcdb.ucla.edu 
}

\begin{linenumbers}
\begin{center}\rule{0.9\linewidth}{0.5pt}\end{center}

Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging 
associated biology. Despite their widespread use, these methods do not account for other factors that also 
contribute to the variability of DNA methylation data. For example, many CpG sites show strong sex-specific 
or cell type specific patterns that likely impact the predictions of epigenetic age.  To overcome these 
limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic 
Pacemaker (MSEPM). We show that the MSEPM is capable of accurately modeling multiple methylation associated factors 
simultaneously, while also providing site specific models that describe the per site relationship between methylation 
and these factors. We utilized the MSEPM with a large aggregate cohort of blood methylation data to construct models 
of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models capture a 
large faction of the variability at thousands of DNA methylation sites.  Moreover, we found modeled sites that are 
primarily affected by aging and no other factors.  Among these, those that lose methylation over time are enriched 
for CTCF transcription factor chip peaks, while those that gain methylation over time are enriched for REST 
transcription factor chip peaks. Both transcription factors are associated with transcriptional maintenance and 
suggest a general dysregulation of transcription with age that is not impacted by sex or cell type heterogeneity. 
In conclusion, the MSEPM is capable of accurately modeling multiple methylation associated factors and the models 
produced can illuminate site specific combinations of factors that affect methylation dynamics.

\begin{center}\rule{0.9\linewidth}{0.5pt}\end{center}
\section{Introduction}

DNA methylation, the addition of a methyl group to the fifth carbon of the cytosine pyrimidine ring,  is 
associated with the topological organization of the cellular genome, gene expression and the state of 
a cell. Within a population of cells the methylation pattern at certain sites can change predictably 
with the age of the individual from which the cells are drawn. This predictable nature of DNA 
methylation has led to the development of accurate DNA methylation based predictive models for 
age and health, termed epigenetic clocks. The difference between the predicted and the expected 
epigenetic age given an individual's chronological age has been interpreted as a measure of age 
acceleration\cite{Horvath2018-ia}, and has been associated with mortality\cite{Perna2016-pi,Marioni2015-sn} 
and other adverse health outcomes\cite{Dugue2018-ad,Huang2019-hf,Armstrong2017-vg,Chuang2017-nk,Horvath2015-af}.

However, epigenetic clocks suffer from several limitations that limit the interpretability of their 
predictions and the underlying mechanisms. Epigenetic clocks are generally trained by using penalized regression 
based methods that attempt to minimize the difference between the predicted and observed value of  age. As a result, 
as the error between predicted and observed age is decreased, the associations between age acceleration and mortality 
disappears\cite{Zhang2019-br}. Second generation epigenetic clocks attempt to resolve this issue by fitting a measure 
of human health, rather than age, and as a result these clocks are generally more sensitive to individual health 
status \cite{Lu2019-lg,Levine2018-en,Belsky2020-ha}. However, while the response variable is modified in these 
clocks the method used to fit the clock is largely the same. Epigenetic clocks are generally trained using 
regularized regression models, where the likelihood is maximized by minimizing the difference between the 
observed and predicted response variable subject to the elastic net penalty,$\lambda_1$ and $\lambda_2$. 
Methylation sites that increase model error and are influenced by other relevant factors such as smoking or obesity, 
may be discarded during model fitting, thus limiting the ability of this approach to account for the effects of these 
extraneous factors on epigenetic aging.

@@

To overcome this limitation, we have developed a multidimensional extension of the EPM, the 
Multi-State Epigenetic Pacemaker (MSEPM).  We show that the (MSEPM) can accurately model  site specific 
methylation variation driven by several factors, and given a trained model, accurately predict the values of the 
factors associated with an individual's observed methylation profile in both simulated methylation datasets and a 
large aggregate blood tissue methylation dataset. Importantly, as factors that explain the observed methylation 
profile of an individual are added to the model the ability to model the factors and methylation values improves. 
Additionally, we show that sites with similar associations to modeled factors cluster together and are enriched 
for specific transcription factors.

\section{Results}

@@

The MSEPM attempts to describe the observed methylation status at any single methylation site as a linear combination 
of methylation associated factors specific to an individual, termed epigenetic factors. Under this model epigenetic 
factors are related to observable individual factors, such as chronological age, sex and cell types, but may be 
transformed relative to observable factors. The epigenetic age factor, for example, often has a non-linear 
relationship with the observed age \cite{Snir2019-ii}. The MSEPM learns the appropriate transformation during model 
fitting to describe the observed methylation status linearly in terms of the epigenetic age factor 
(and not linearly with age). Given a site $i$ and individual $j$ the observed methylation status can be 
modeled as $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} f_{j,k} r_{i,k}$, where $k$ epigenetic factors, $f_{j,k}$, 
are weighted by $k$ site specific parameters, $r_{i,k}$, and offset by a sites specific intercept term, $r^0_i$. 
Site parameters, $r_{i,k}$ and $r^0_i$, are characteristic of the site and shared amongst all individuals while 
epigenetic factors, $f_{j,k}$, are characteristic of an individual and are the same across all sites for that 
individual. As a consequence, observed methylation differences between any two individuals are dependent on 
individual epigenetic factors $f_{j,k}$. In practice, the observed methylation value is also dependent on a 
normally distributed error term 
$\epsilon_{i,j}$, $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} s_{j,n} r_{i,n} + \epsilon_{i,j}$.

Given an input matrix $\hat{M} = [m_{i,j}]$ the objective of the MSEPM is to find the optimal values of 
$r_{i,k}$ and $f_{k,j}$ that minimize the  residual sum of square (RSS) error, 
$\epsilon_{i,j}^2 = (\hat{m_{i,j} – r^0_{i} - \sum^n_{n=1} f_{j,k} r_{i,k} })^2$. This is 
accomplished through the implementation of a conditional expectation maximization algorithm. The 
maximum likelihood (ML) values of $r_{i,k}$ and $r^0_i$ for each $i$ methylation site can be solved 
using ordinary least squares (OLS) regression. Given the ML $r_{i,k}$ site coefficients each $k$ epigenetic 
factor is then updated by fixing the site coefficients and updating $s_{k,j}$ by minimizing the RSS across 
all $i$ sites using gradient descent, $f_{k,j}^{n+1} = f_{k,j} - \lambda \nabla F(f_{k,j})$, where $\lambda$ 
is a specified learning rate. The optimization is accomplished by alternating between optimizing $r_{k,i}$ and 
$f_{k,j}$ until the reduction in sum of the site RSS is below a specified threshold between iterations or a set 
number of interactions is reached. Importantly, while the ML values of $f_{j,k}$ are by definition linear with 
the methylation status at any site, the original input factors for $f_{j,k}$ may not be.

Provided a trained MSEPM model and an unobserved methylation matrix, epigenetic factors are estimated by 
calculating each independent OLS for solution all $i$ sites given the $r_{i,k}$ coefficients set for the 
respective input factor.  These epigenetic factors can then be used to find the expected methylation value using the 
trained individual site models where $E[m_{i,j}] = r_{i,0} + F_{j} \dot R_{i}$.

\subsection{Simulated Methylation Associated Factors}

We implemented a simulation framework that extends the MSEPM model where the methylation status at site $i$ for an 
individual $j$ is described by a weighted sum of epigenetic factors, 
$\hat{m_{i,j}} = r^0_i + \sum^n_{l=1} f_{k,j} r_{i,k} + \epsilon_{i,j}$. As shown in our 
previous work \cite{Snir2019-ii} the association between a methylation associated input factor and methylation 
status is not necessarily linear. To account for non-linear associations, $f_{k,j}$, have the 
form $f_{j,k} =  q_{j,k}^{\gamma_{k}}$, where $\gamma_{k}$ is the transformation between a factor of 
magnitude $q_{n,j}$ and the epigenetic factor. In practice the value of the $f_{k,j}$ is often unknown and the 
association between methylation status and $f_{k,j}$ is inferred through $q_{n,j}$. We simulated individuals 
whose methylation is determined by four factors and their associated epigenetic factors: a uniform factor 
approximating age with a non-linear association with methylation status 
($q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}$, Figure 1A-B), a binary trait resembling sex, 
linearly associated with methylation status ($q \sim B(1,.5), s_{Sex}=q$, Figure 1C-D), a continuous normal 
(CN) phenotype resembling a cell type with a linear association with methylation 
status $(q \sim\mathcal{N}(1, 0.1), s_{CN}=q$, Figure 1E-F), and a continuous exponentially (CE) distributed trait 
resembling obesity with a linear association with methylation status 
$(q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, Figure G-H)$. Methylation sites were simulated by first randomly 
setting the dynamic range of the methylation site, ($ -1< \delta < 1$), a site intercept, $r^0_i$, and the 
site error, $(\sigma_i \sim \mathcal{U}(0.025, 0.05)$). Simulated methylation sites were associated with a 
combination of zero, one, or multiple epigenetic factors. Rates for sites associated with multiple factors were 
set by sampling from a uniform distribution. Site rates were normalized to describe the dynamic range of the 
simulated methylation site unless the rates for all phenotypes were zero.

We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. 
We simulated 1000 samples with the four epigenetic factors described above. We then simulated methylation values
 using the simulated site rates. Simulated samples were then split for training ($n=500$) and testing ($n=500$).
  MSEPM models were then fitted using the values of the input factors, $q_{k,j}$, as the initial guess for
   $f_{k,j}$. We generated 1000 simulated datasets and fit MSEPM models using four combinations of input 
   factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). Within each simulation, epigenetic state predictions and 
   methylation site predictions were made for all testing samples. All models captured the nonlinear association 
   between simulated age and methylation (S. Figure 1). As the number of factors in the model is increased the mean 
   absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases 
   (Figure 2A). Importantly, to accurately assess simulated age it is necessary to account for the influence of 
   the other simulated factors (Sex, CN, CE). The MAE between the predicted and simulated methylation values decreases 
   as simulated factors are added to the model, and accurately assessing the methylation status of a simulated site 
   requires that the factor associated with the methylation status at the site is included in the model (Figure 2A).

The MSEPM model generated using all four simulated factors can capture the relative magnitude of the simulated 
site-specific rates (Figure 2C-F).  However, the model has difficulty capturing the exact relationship between the 
simulated factors (age, CN and CE) and the inferred factors (Figure 2C, E-F). This is likely due to limitations of 
the model at capturing nonlinear methylation association and a limited training range for normally and exponentially 
distributed traits. Regardless, the four-factor model can accurately predict the simulated methylation value 
(Figure 2 D) and site intercept (Supp. Figure 1G). We also assessed the model robustness to variation in the 
number of samples and sites used for model training by randomly selecting a reduced subset of samples or sites for 
model training. MSEPM models trained with age, sex, CN, and CE can accurately assess all simulated phenotypes with 
few samples and sites (Supp Figure2 B-E).

\subsection{Blood MSEPM Model}

We utilized a large aggregated dataset composed of Illumina 450k array data from 17 publicly available 
datasets \cite{Ventham2016-qj,Demetriou2013-wb,Polidoro2013-xv,Johansson2013-of,Arloth2020-lo,Liu2013-dg,Soriano-Tarraga2016-uq,Chuang2017-nk,Zannas2019-me,Kurushima2019-pe,Voisin2015-lh,Tan2014-sg,Tserel2015-ro,Butcher2017-oz,Dabin2020-iw,Marabita2013-cj,ValleG2019-xi} 
deposited in the Gene Expression Omnibus \cite{Barrett2012-gu} (GEO) generated from blood derived samples 
(whole blood, peripheral blood lymphocytes, and peripheral blood mononuclear cells). All datasets were processed 
through a unified pipeline from raw array intensity data (IDAT) files using minfi (Aryee et al., 2014). Sex and 
blood cell type abundance predictions were made for each as previously described\cite{Houseman2012-rr,Aryee2014-ky}. 
The aggregate data spanned a wide age range (0.0  - 99.0 years, Figure 3A), contained more predicted females 
($n=3392$) than males ($n=2295$, Figure 3B) and produced reasonable cell type abundance estimates (Figure 3C).

We trained MSEPM models using data assembled from four GEO 
series\cite{Johansson2013-of,Liu2013-dg,Butcher2017-oz,Damaso2020-gd}  ($n=1605$) with samples spanning a wide age 
range (0.01 - 94.0 years). The samples were randomly split into training ($n=1203$) and validation ($n=402$) sets 
stratified by age. Training set blood cell type abundance estimates were used to train a principal component analysis 
(PCA) model which was then used to calculate cell type PCA estimates for the validation and testing sets. The first 
cell type principal component (CT-PC1) is largely driven by the relative abundance of granulocytes (Figure 3D), while 
the second PC (CT-PC2) captures relative differences in the abundance of differentiated lymphocytes (Figure 3D).  
Methylation values for all samples were quantile normalized by probe type \cite{Horvath2013-sk} using the median 
site methylation values across all training samples for each methylation site.

Methylation sites were selected for modeling with MSEPM if the sites were highly correlated with 
age ($n=276$) , sex ($n=49$), CT-PC1 ($n=120$), CT-PC2 ($n=116$) or a combination of factors ($n=238$), 
a total of 778 unique sites. Four MSEPM models were fit using the selected sites with four combinations of input 
factors (Age, Age Sex, Age Sex CT-PC1, and Age Sex CT-PC1 CT-PC2). Training sample factors were min-max scaled between 
0 and 1 before model training. The association between the fit epigenetic factor predictions against the input modeled 
factors was assessed by fitting a trendline between epigenetic state predictions and scaled continuous input factors 
using the state prediction made for the MSEPM model trained with all four input factors. Performance of the MSEPM 
model was then evaluated using the testing samples ($n=4,082$). The performance of the MSEPM largely closely 
resembles the simulation results. All four MSEPM models capture the nonlinear relationship between age and 
methylation status (Supp. Figure 6). The epigenetic state prediction associated with age improves as the underlying 
methylation data are more fully explained through the addition of epigenetic factors (Supp. Figure 6). The MSEPM model 
fit with Age, Sex, CT-PC1 and CT-PC2 can accurately model the associated epigenetic state for each factor (Figure 4 A-D) 
and accurately predicts the methylation levels at individual sites ($R^2=0.935$, $MAE=0.035$, Figure 4 E). 
The trained MSEPM produces a collection of methylation site models that can help explain the association between 
modeled factors and methylation status.

\subsection{Analysis of chromatin regulators of site clusters}

We evaluated the relationship between sites that are influenced by age, sex, CT-PC1 or CT-PC2  and potential 
regulatory factors by performing overlap enrichment analysis of these sites with transcription factor chromatin 
immunoprecipitation peaks present in the ENCODE V4 \cite{ENCODE_Project_Consortium2012-oe,Davis2018-ha} release. We 
first identified sites with similar coefficients of epigenetic factors through hierarchical clustering. Briefly, 
sites were clustered using Ward's method by the euclidean distance between site regression coefficients that were 
normalized by the standard deviation of the methylation values amongst the training samples. The resulting tree was 
cut at a height of 18 to produce 10 distinct clusters with clear associations to the modeled factors (Figure 5A).

>>>>

\large{\bf{The Multi-State Epigenetic Pacemaker enables the 
identification of combinations of factors that influence DNA methylation}}\\ \\
\noindent{\small Colin Farrell$^{1,4}$, Keshiv Tandon$^{1}$, Roberto Ferrari$^{2}$, Kalsuda Lapborisuth$^{1}$, 
                 Sagi Snir$^{3}$,\\ and Matteo Pellegrini$^{1,4}$\\ \\
\noindent{\footnotesize 
$^1$Dept. of Molecular, Cell and Developmental Biology; \\ University of
California, Los Angeles, CA 90095, USA;;\\
\noindent
$^2$Dept. of Chemistry, Life Sciences and Environmental Sustainability, Laboratory of Molecular Cell Biology of the Epigenome (MCBE), University of Parma, Italy;\\
$^3$Dept. of Evolutionary Biology, University of  Haifa, Israel;\\
$^4$Corresponding Authors; colinpfarrell@gmail.com, matteop@mcdb.ucla.edu 
}

\begin{linenumbers}

\abstract{Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
Despite their widespread use, these methods do not account for other factors that also contribute to the variability of DNA methylation data.
 For example, many CpG sites show strong sex-specific or cell type specific patterns that likely impact the predictions of epigenetic age.  
 To overcome these limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic Pacemaker (MSEPM). 
 We show that the MSEPM is capable of accurately modeling multiple methylation associated factors simultaneously, while also providing site specific 
 models that describe the per site relationship between methylation and these factors. We utilized the MSEPM with a large aggregate cohort of blood
  methylation data to construct models of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models 
  capture a large faction of the variability at thousands of DNA methylation sites.  Moreover, this approach allows us to identify sites that 
  are primarily affected by aging and no other factors.  An analysis of these sites reveals that those that lose methylation over time are 
  enriched for CTCF transcription factor chip peaks, while those that gain methylation over time are associated with bivalent promoters of 
  genes that are not expressed in blood. These observation suggest mechanisms that underlie age associated methylation changes and suggest 
  that age associated increases in methylation may not have strong functional consequences on cell states. In conclusion, the MSEPM is 
  capable of accurately modeling multiple methylation associated factors and the models produced can illuminate site specific 
  combinations of factors that affect methylation dynamics.}

\section{Introduction}\label{sec1}

DNA methylation, the addition of a methyl group to the fifth carbon of the cytosine pyrimidine ring,  is associated with 
the topological organization of the cellular genome, gene expression and the state of a cell. Within a population of cells 
the methylation pattern at certain sites can change predictably with the age of the individual from which the cells are 
drawn. This predictable nature of DNA methylation has led to the development of accurate DNA methylation based predictive 
models for age and health, termed epigenetic clocks. The difference between the predicted and the expected epigenetic 
age given an individual’s chronological age has been interpreted as a measure of age acceleration\cite{Horvath2018-ia}, 
and has been associated with mortality\cite{Perna2016-pi,Marioni2015-sn} and other adverse health 
outcomes\cite{Dugue2018-ad,Huang2019-hf,Armstrong2017-vg,Chuang2017-nk,Horvath2015-af}.

However, epigenetic clocks suffer from several limitations that limit the interpretability of their 
predictions and the underlying mechanisms. Epigenetic clocks are generally trained by using penalized regression based 
methods that attempt to minimize the difference between the predicted and observed value of  age. As a result, as the 
error between predicted and observed age is decreased, the associations between age acceleration and mortality 
disappears\cite{Zhang2019-br}. Second generation epigenetic clocks attempt to resolve this issue by fitting a 
measure of human health, rather than age, and as a result these clocks are generally more sensitive to individual 
health status\cite{Lu2019-lg,Levine2018-en,Belsky2020-ha}. However, while the response variable
is modified in these clocks the method used to fit the clock is largely the same. Epigenetic clocks are generally 
trained using regularized regression models, where the likelihood is maximized by minimizing the difference between 
the observed and predicted response variable subject to the elastic net penalty,$\lambda_1$ and $\lambda_2$. Methylation 
sites that increase model error and are influenced by other relevant factors such as smoking or obesity, may be 
discarded during model fitting, thus limiting the ability of this approach to account for the effects of these 
extraneous factors on epigenetic aging.

@@

To overcome this limitation, we have developed a multidimensional extension of the EPM, the Multi-State 
Epigenetic Pacemaker (MSEPM).  We show that the (MSEPM) can accurately model  site specific methylation 
variation driven by several factors, and given a trained model, accurately predict the values of the 
factors associated with an individual’s observed methylation profile in both simulated methylation 
datasets and a large aggregate blood tissue methylation dataset. Importantly, as factors that explain 
the observed methylation profile of an individual are added to the model the ability to model the factors 
and methylation values improves. Additionally, we show that sites with similar associations to modeled factors 
cluster together and are enriched for specific transcription factors. Therefore, unlike traditional epigenetic 
clocks, the MSEPM allows us to study mechanisms that may underlie age associated methylation changes.  In our 
large dataset of blood samples, we find that sites that increase methylation with age are enriched for bivalent 
promoters, and are proximal to genes that are lowly expressed in blood.  These results suggest that positively 
age associated sites may not have a significant functional impact on aging traits. The MSEPM is available as a 
Python package with scikit-learn style syntax under a MIT license at https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.

\section{Methods}\label{sec2}

@@

The MSEPM model describes the observed methylation at site $i$ and for individual $j$, $\hat{m_{i,j}}$, 
as a weighted linear combination of $k$ individual epigenetic factors $p_{k,j}$. 
$$\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} p_{j,k} r_{i,k}$$ Where $k$ epigenetic factors are weighted by $k$ site specific 
epigenetic rates of change, $r_{i,k}$, and offset by a sites specific intercept term, $r^0_i$. Site parameters,
 $r_{i,k}$ and $r^0_i$, 
are characteristic of the site and shared amongst all individuals while epigenetic factors, $p_{j,k}$, are characteristic 
of an individual and are the same across all sites for that individual. In practice, 
the observed methylation value is also dependent on a normally distributed error term 
$\epsilon_{i,j}$.

$$\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} p_{j,k} r_{i,k} + \epsilon_{i,j}$$

Under this model epigenetic factors are related to observable individual 
factors $p_{k,j}^0$, such as chronological age, sex and cell types, but may be transformed relative to observable factors. 
The epigenetic age factor, for example, often has a non-linear relationship with the observed age\cite{Snir2019-ii}. 
The MSEPM learns the appropriate transformation during model fitting to describe the observed methylation status 
linearly in terms of the epigenetic age factor, but not linearly with age.

Given an input matrix $\hat{M} = [m_{i,j}]$ of methylation values for $i$ sites and matched observable epigenetic factors 
$\hat{P^0} = [p_{j,k}^0]$ for $j$ individuals the objective of the MSEPM is to find the optimal values of 
$r_{i,k}$ and $p_{j,k}$ that minimize the  residual sum of square (RSS) error,

$$\epsilon_{i,j}^2 = (m_{i,j} - r^0_{i} - \sum^n_{k=1} p_{j,k} r_{i,k})^2$$

This is accomplished through the implementation of a conditional expectation maximization algorithm. The maximum likelihood (ML) values of 
$r_{i,k}$ and $r^0_i$ can
be solved using ordinary least squares (OLS) regression. Provided the ML estimates for $r_{i,k}$, the site coefficients are fixed and 
epigenetic factors, $p_{j,k}$, are  updated by minimizing the RSS across all $i$ sites using 
gradient descent, $$p_{j,k}^{n+1} = p_{j,k}^n - \lambda \nabla F(p_{j,k})$$ where $\lambda$ is a specified learning rate. 
The optimization is accomplished by alternating between optimizing $r_{i,K}$ 
and $p_{j,k}$ until the reduction in sum of the site RSS is below a specified threshold or a set number 
of iterations is reached. Importantly, while the ML values of $p_{j,k}$ are 
by definition linear with the methylation status at any site, the original input factors for $p_{j,k}^0$ may not be.

Provided a trained MSEPM model and an unobserved methylation matrix, epigenetic factors are estimated by calculating each independent OLS for solution all $i$ 
sites given the $r_{i,k}$ coefficients set for the respective input factor.  These epigenetic factors can then be used to find the expected methylation value 
using the trained individual site models where $$E[m_{i,j}] = r_{i,0} + P_{j} \dot R_{i}$$

\subsection{MSEPM Simulation Framework}

We implemented a simulation framework using the MSEPM formulation evaluate the performance of the MSEPM model under various conditions.
To simulate the association between methylation status and an observable factor we modeled the 
epigenetic factor $p_{k,j}$ as function of time, or age, and magnitude, $p_{k,j}^0$ with a non-linear transformation $\gamma_{k}$, 
where $p_{k,j} = Age_j p_{k,j}^{0,\gamma_{k}}$. In practice the value of the $p_{k,j}$ is often unknown and the association between 
methylation status and $p_{k,j}$ is inferred through the observable factor $p_{k,j}^0$.

Methylation sites were simulated by first randomly setting the dynamic range of the methylation site, 
$-1< \delta < 1$ a site intercept, $r^0_i$, and the site error, $\sigma_i \sim \mathcal{U}(0.025, 0.05)$. The dynamic 
range of the methylation site is described by the initial methylation value, $m_0 ~ \beta(.2,.2)$, 
and the target methylation value, $m_t$, where the dynamic range is $\delta_i = m_t - m_o$.
The value of $m_t$ is set conditionally to ensure dynamic range is always larger than some specified threshold, $\theta$, where 
$ \theta \leq|\delta| \geq. r^0_i ~ \beta(.2,.2)$.

Simulated methylation sites are then randomly associated with a combination of zero, one, or multiple epigenetic factors. Rates for sites 
associated with multiple factors were set by sampling from a uniform distribution. The weighted factor rates are normalized so 
the input combination of traits describes the dynamic range of the simulated site, $\delta$. If a site is associated 
with no factors the observed methylation status of a site is described by a random normal with a characteristic offset, $\hat{m_i} = r_{0,i} + N(\mu, \sigma)$.

\subsection{Blood MSEPM Model Training}
MSEPM models were trained using a large aggregate dataset of blood derived methylation data from 17 publicly available 
datasets\cite{Ventham2016-qj,Demetriou2013-wb,Polidoro2013-xv,Johansson2013-of,Arloth2020-lo,Liu2013-dg,Soriano-Tarraga2016-uq,Chuang2017-nk,
Zannas2019-me,Kurushima2019-pe,Voisin2015-lh,Tan2014-sg,Tserel2015-ro,Butcher2017-oz,Dabin2020-iw,Marabita2013-cj,ValleG2019-xi}. 
Illumina methylation 450K Beadchip methylation array IDAT files were processed using minfi\cite{Aryee2014-ky} (v1.34.0). 
Sample IDAT files were processed in batches according to GEO series and Beadchip  identification. Methylation values within 
each batch were normal-exponential normalized using out-of-band probes\cite{Triche2013-pp}. Blood cell types counts were estimated 
using a regression calibration approach\cite{Houseman2012-rr} and sex predictions were made using the median intensity measurements
of the X and Y chromosomes as implemented in minfi\cite{Aryee2014-ky}. Samples were filtered for quality control using the the relative 
intensity of the methylated and unmethylated probes. Samples were used for downstream analysis if the sample median 
methylation probe intensity was greater than 10.5 and the difference between the observed and expected median 
unmethylation probe intensity is less than 0.4, where the expected median unmethylated intensity is described by 
$E[intensity_{unmethylated}] = 0.66 intensity_{methylated} + 3.718$. This resulted in a total of 5687 samples.

We trained MSEPM models using data assembled from four GEO series\cite{Johansson2013-of,Liu2013-dg,Butcher2017-oz,Damaso2020-gd}  ($n=1605$). 
The samples were randomly split into training ($n=1203$) and validation ($n=402$) sets stratified by age. Methylation values for all 
samples were quantile normalized by probe type\cite{Horvath2013-sk} using the median site methylation values across all training samples for each methylation site.
Training set blood cell type abundance estimates were used to train a principal component analysis (PCA) model which was then used to calculate cell 
type PCA estimates for the validation and testing sets. Methylation sites were selected for modeling with MSEPM if the site methylation values were correlated with age ($n=276$) , 
sex ($n=49$), CT-PC1 ($n=120$), CT-PC2 ($n=116$) or a combination of factors ($n=238$) by absolute pearson correlation coefficient. Where a absolute pearson correlation coefficient greater than 0
.7, 0.995, 0.92 and 0.64 for age, sex, CT-PC1 and CT-PC2 respectively. Sites with a sum of absolute pearson coefficients across the four factors greater than 1.8 were also included $(n=238)$ 
for a total of 778 methylation sites. Min-max, (0-1), scalers were fit using the training input features. Validation and testing sample features were transformed with the trained scalers. Age was min-max 
scaled on a range from 0-100 years. MSEPM models were trained with a learning rate of 0.01 with an iteration limit of 200.

\subsection{Blood MSEPM Model Cluster Transcription Factor Overlap Analysis}
We evaluated the relationship between modeled sites, input factors and regulatory transcription factors using overlap enrichment analysis.  
We built a custom transcription factor reference set using ENCODE V4 transcription factor chromatin immunoprecipitation\cite{ENCODE_Project_Consortium2012-oe,Davis2018-ha} 
(release 1.4.0 - 2.1.2)  irreproducible discovery rate narrow bed peaks, 
  which contains peaks with high rank consistency between replicates, that were not audited for non-compliance or errors. 
  GRCh38 region coordinates were lifted to GRCh37 coordinates using liftOver\cite{Hinrichs2006-oq}. The overlap reference 
  contains 714 transcription factor targets from 1621 accession IDs.

We then performed hierarchical clustering of the four factor MSEPM model sites based on the similarity of their regression coefficients.
Individual methylation site coefficients were first normalized by the standard deviation of methylation values of the site among the training samples, $r_{i,k} / \sigma_{i}$. 
A distance matrix was then created by taking the Euclidean distance between the normalized site model coefficients. Sites were then clustered using Ward’s 
method which seeks to minimize within cluster variance by minimizing the increase in the error sum of squares (ESS) through successive cluster fusions.  
Clusters label by tree cutting at a height of 18. All clustering analysis was carried out using SciPy v1.6.3\cite{Virtanen2020-mm}.

Transcription factor enrichment analysis was performed with  LOLA\cite{Sheffield2016-wg} which assesses the genomic 
region set overlap between a set of query regions and a set of reference regions, within a specified shared background set, 
using Fisher's exact test. Overlap analysis was performed for sites within a cluster against the ENCODE V4 reference region 
(1BP minimum overlap) using all sites assayed with Infinium HumanMethylation450K BeadChip as background.

\subsection{Clustering sites with age-associated increases in methylation}
To better understand age associated methylation in whole blood, we examined each site within MSEPM four factor blood model cluster 7 individually, 
as this cluster contains sites that have methylation that increases with age but is not strongly affected by other factors. 
Using the EWAS Data Hub (Xiong, et al. 2016), we validated our results by obtaining additional methylation by age data in 
whole blood for each site in the cluster (McCartney, et al. 2019).  We created a matrix with every sample and its associated 
methylation and age from cluster 7, then used age associated methylation levels to create a clustered heatmap using the Matlab 
function Clustergram. We then clustered the tree into four groups which were analyzed separately.

We also identified the genes that were proximal to each site using Cistrome-GO (Li et al. 2019). We then 
examined the expression of the genes across tissues in the Genotype-Tissue Expression (GTEx) database database. 
We used the GTEx Multi Gene Query to find which tissues those genes belonged to.

We utilized the Toolkit for Cistrome Data Browser \cite{Zheng2019-vm,Mei2017-yq} for the analysis of significant factors in each cluster. This allowed 
us to input .bed files of each sub-cluster and generate a GIGGLE score for specific transcription factors, histone marks, 
and chromatin regions to assess significance of these elements. A GIGGLE score tailored ranking of loci based on overlap 
of genomic features provided by the user\cite{Layer2018-gr}.

\subsection{H3K4me3 enrichment analysis} 
Enrichment of analysis for H3K4me3 (figure 7A) was carried out by downloading rpm normalized bigwig files of H3K4me3 ChIP-seq data from 
epigenomesportal\cite{Bujold2016-vk} for CD38+ B Cells and CD56+ NTK Cells (for both 0-5 years old and 60-65 years old individuals). 
Heatmaps of H3K4me3 were generated using deepTools2\cite{Ramirez2016-xl} using the computeMatrix and plotHeatmap function to plot the 
bigwig signal over genomic regions of cluster 7 as the BED input. The IGV genome browser\cite{Robinson2011-he} was used to generate an image of the KCTD1 
and IRS2 promoter regions shown in figure 7B using downloaded bigwig tracks.

\subsection{Analysis Environment}
Analysis was carried out in a Jupyter\cite{Basu_undated-vq} analysis environment. 
Joblib\cite{Varoquaux2009-al}, SciPy\cite{Virtanen2020-wt}, Matplotlib\cite{Hunter2007-nq}, 
Seaborn\cite{Waskom2021-gj}, Pandas\cite{McKinney2012-ta} and TQDM\cite{Da_Costa-Luis2019-lr} 
packages were utilized during analysis.

\section{Results}\label{sec3}

\subsection{Simulated Methylation Associated Traits}

We simulated individuals whose methylation is determined by four factors and their associated epigenetic factors: 
a uniform factor approximating age with a non-linear association with methylation status 
$$q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}, \text{Figure 1A-B}$$ a binary trait resembling sex, linearly associated 
with methylation status $$q \sim B(1,.5), s_{Sex}=q, \text{Figure 1C-D}$$ a continuous normal (CN) phenotype 
resembling a cell type with a linear association with methylation status $$q \sim\mathcal{N}(1, 0.1), s_{CN}=q, 
\text{Figure 1E-F}$$ and a 
continuous exponentially (CE) distributed trait resembling obesity with a linear association with methylation status 
$$q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, \text{Figure 1G-H})$$

We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. We simulated 1000 samples with the 
four epigenetic factors described above. We then simulated methylation values using the simulated site rates. Simulated samples
 were then split for training ($n=500$) and testing ($n=500$). MSEPM models were then fitted using the values of the input factors, $p^0_{k,j}$. 
 We generated 1000 simulated datasets and fit MSEPM models using four combinations of input factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). 
 Within each simulation, epigenetic state predictions and methylation site predictions were made for all testing samples. All models captured 
 the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased the mean 
 absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE). The MAE between the predicted 
  and simulated methylation values decreases as simulated factors are added to the model, and accurately assessing the methylation status of a 
  simulated site requires that the factor associated with the methylation status at the site is included in the model (Figure 2A).
